Reasoning about Connectors Using Coq and Z3

### 2018-08-12 FACS_Extension/paper_revised0508.tex
minor issues corrected.

\newcommand{\redt}[1]{\textcolor{red}{#1}}
%\newcommand{\liyi}[1]{\textcolor{blue}{#1}}
%\newcommand{\xy}[1]{\textcolor{blue}{#1}}

@@

%\maketitle
\begin{abstract}
Reo is a channel-based exogenous coordination language in which complex coordinators, called connectors, are compositionally built out of simpler ones. In this paper, we present an approach to model and reason about connectors using Coq and Z3.
The model reflects the original structure of connectors simply and clearly. In our framework, both basic connectors (channels) and composition operations are interpreted as axioms. Furthermore, connectors are interpreted as logical predicates which describe the relations
between inputs and outputs. With such definitions provided, connector properties, as well as equivalence and refinement relations between different connectors, can be naturally formalized as \emph{goals} in Coq and proved using pre-defined \emph{tactics}, if satisfied by connectors. Furthermore, when fail to prove the properties, we use Z3, the SMT solver, to search for possible bounded counter examples automatically.
% Particularly, to fulfill the proof of refinement relation, Z3 is used as an assistant tool for solving logical formulas.
%\paragraph*{Keywords:}

@@

Modern software systems are typically distributed over large networks of computing devices, and usually the components
that comprise a system do not exactly fit together as pieces of a jigsaw puzzle, but leave significant interfacing gaps
that must somehow be filled with additional code. Compositional coordination models and languages provide a formalization
of the ``glue code" that interconnects the constituent components and organizes the mutual interactions among them in a
distributed processing environment, and played a crucial role for the success of component-based systems in the past decades.

As an example, Reo \cite{Arb04}, which is a channel-based exogenous coordination language, offers
a powerful framework for implementation of coordinating component connectors. Connectors provide the protocols that control and organize
the communication, synchronization and cooperation among the components that they interconnect. Primitive connectors, called {\em channels}
in Reo, can be composed to build complex connectors. Reo has been successfully applied in different application domains, such as
service-oriented computing and bioinformatics \cite{CCA04, SA07}. In recent years,
verifying the correctness of connectors is becoming a critical challenge, especially due to the advent of Cloud computing
technologies. The rapid growth of size and complexity of the computing infrastructures has made it more difficult to model
and verify connector properties, and thus leads to less confidence on the correctness of connectors.

Several works have been done for formal modeling and verifying
connectors. An operational semantics for Reo using Constraint Automata
(CA) was provided by Baier et al. \cite{BSAR06}, and later the symbolic model checker Vereofy \cite{BBK+10} was developed,
which can be used to check CTL-like properties.
% However, modeling unbounded primitives or even bounded primitives with unbounded
% data domains is impossible with finite constraint automata. Bounded large data domains cause an
% explosion in the constraint automata model which becomes problematic.
% A model for Reo connectors based on the idea of coloring a connector with possible data flows to
% resolve synchronization and exclusion constraints was presented by Clarke et al. \cite{CCA07}. Unlike the operational semantics,
% data sensitive behavior, which is supported by Reo, are not captured by the coloring approach.
Besides, one attractive approach is to translate from Reo to other formal models such as Alloy
\cite{KSA+08}, mCRL2 \cite{KKV12}, UTP \cite{AAA+09,SAA+12}, etc.,
which makes it possible to take advantage of existing verification
tools. A comparison of existing semantic models for Reo can be found
in \cite{JA12}.

In a previous paper \cite{ZHL+17} we showed how to formally model and reason about connectors' properties in Coq. The basic idea of our approach is to model the behavior of a connector by representing it as a logical predicate which describes the relation among the timed data streams on the input and output nodes, and to reason about connectors' properties, as well as the equivalence and refinement relations between connectors, by using proof principles and tactics in Coq. Compared with existing approaches for verifying connectors' properties \cite{BBK+10,KB09,KKV12}, reasoning about connectors using theorem provers like Coq is especially helpful when we take infinite behavior into consideration. The coinductive proof principle makes it possible to prove connectors' properties easily while it is difficult (sometimes impossible) for other approaches (like model checking) because of the huge (or maybe infinite) number of states.

In the current paper, we aim to provide a more complete account of the approach in \cite{ZHL+17} to formally model and reason about connectors. The main drawback of failing to check the complementary refinement relation in Coq is partially made up by using the constraint solver Z3. While Coq provides a platform to prove properties with a series of tactics and strategies to be considered, Z3 can be used as a component in the context %of other tools
that requires solving logical formulas. The logical formulas involved in the model of Reo should reflect the constraints of connectors and be consistent with their behavior.

The approach is not a brand new idea, as we have already provided a solution for modeling Reo in Coq in \cite{LS15}, where connectors are represented in a constructive way, and verification is essentially based on simulations. We do believe that the approach in this paper is reasonably different from its predecessor \cite{LS15} where Coq seldom shows its real power. To be more specific, our new work has its certain
advantages comparing with \cite{LS15} in the following aspects:
\begin{itemize}
\item {\bf Modeling Method:} We use axioms to describe basic channels and their composition operations, which is more natural on a proof-assistant platform than the simulation-based approach in \cite{LS15}.
\item {\bf Expression Power:} Any valid Coq expression can be used to depict properties, which is obviously more powerful than just using LTL formulas in \cite{LS15}. Furthermore, support for continuous time behavior is also possible in our approach in this paper.
\item {\bf Refinement and Equivalence Checking:} In our earlier work \cite{ZHL+17}, equivalence and refinement relations can be proved among different connectors, while the approach in \cite{LS15} is not capable of either equivalence or refinement checking. Furthermore, the constraint solver Z3 is adopted in this work to search for bounded counter examples when the refinement relations cannot be proved in Coq.
\end{itemize}

The paper is organized as follows. After this general introduction, we briefly summarize Reo, Coq and Z3
in Section \ref{sec:pre}. Section \ref{sec:channelandoperator} introduces the
formal model of basic channels, operators and complex connectors in Coq. Section
\ref{sec:verification} shows how to reason about connector properties and equivalence (or
refinement) relations under our Coq formalization. Section \ref{sec:refinement} provides a Reo model in Z3 and shows the usage of Z3 in refinement checking.
In Section \ref{sec:conclusion}, we conclude with some further research directions. Full source codes can be found at \cite{reo2coq2Z3} for further reference.

@@

In this section, we provide a brief introduction to the coordination language Reo, and the proof assistant tools Coq and Z3.

\subsection{The Coordination Language Reo}
Reo is a channel-based exogenous coordination language wherein complex coordinators, called connectors,
are compositionally built out of simpler ones \cite{Arb04}.
Further details about Reo and its semantics can be found in \cite{Arb04, AR03, BSAR06}.
The simplest connectors are channels with well-defined behavior such as synchronous channels, FIFO channels, etc.
% Each channel in Reo has exactly two directed ends, with their own identities.
There are two types of channel ends: \emph{source} ends and \emph{sink} ends. A source channel end accepts
data into the channel, and a sink channel end dispenses data out of the channel.
\begin{figure}
  \centering
  \begin{tikzpicture}
  \draw (0.7, 0.6) node[draw=none] {Sync};
  \sync{(0,1)}{(1.5,1)}{}
  \draw (2.7, 0.6) node[draw=none] {SyncDrain};
  \syncdrain{(2,1)}{(3.5,1)}{node[below]{}}
  \draw (4.7, 0.6) node[draw=none] {FIFO1};
  \fifoe{(4,1)}{(5.5,1)}{node[below]{}}
  \draw (6.7, 0.6) node[draw=none] {LossySync};
  \lossysync{(6,1)}{(7.5,1)}{node[below left]{}}
  \end{tikzpicture}\vspace{-6mm}
  \caption{Four types of basic channels.}\label{fig:basicchannel}
\end{figure}

The graphical notations of some basic channels are presented in Figure~\ref{fig:basicchannel}, and their behavior can be interpreted as follows:
\begin{itemize}
\item{\textbf{Sync:} a synchronous channel with one source end and one sink end. The pair of I/O operations on its two ends can succeed only simultaneously.}
\item{\textbf{SyncDrain:} a synchronous channel which has two source ends.
 The pair of input operations on its two ends can succeed only simultaneously.
 All data items written to this channel are lost.}
\item{\textbf{FIFO$n$:} an asynchronous channel with one source end and one sink end, and a bounded buffer with capacity $n$.
It can accept data items from its source end. The accepted data items are kept in the internal buffer, and dispensed to
the sink end in FIFO order.
Especially, the FIFO$1$ channel is an instace of FIFO$n$ where the buffer capacity is 1.}
%\item{\textbf{AsyncDrain:}  an asynchronous channel} which has two source ends. The channel guarantees that the operations on its two ends never succeed simultaneously. All data items written to this channel are lost.
\item{\textbf{LossySync:} a synchronous channel with one source end
    and one sink end. The source end always accepts all data items. If
    there is no matching output operation on the sink end of the
    channel at the time that a data item is accepted, then the data
    item is lost; otherwise, the channel transfers the data item
    exactly the same as a Sync channel, and the output operation at the sink end succeeds.}
\end{itemize}

@@

Complex connectors are constructed by composing simpler ones via the \emph{join} and \emph{hiding} operations. Channels are joined together in nodes. The set of channel ends coincident on a node is disjointly partitioned into the sets of source and sink channel ends that coincide on the node, respectively. Nodes are categorized into source, sink and mixed nodes, depending on whether all channel ends that coincide on a node are source ends, sink ends or a combination of the two. The hiding operation is used to hide the internal topology of a connector. The hidden nodes can no longer be accessed or observed from outside. There are three types of operations for channel composition: \emph{flow-through}, \emph{merge} and \emph{replicate}. Figure~\ref{fig:channelcomposition} provides the graphical representation of these operations.

@@

Coq \cite{huet1997coq} is a widely-used proof assistant tool, where denotational formalizations (e.g.
theorem and hypothesis) and operational formalizations (e.g. functions and algorithms) are naturally
integrated. Moreover, it allows the interactive construction of formal proofs.
The formal language used in Coq, \emph{Gallina}, provides a convenient way to define
both programming statements and mathematical propositions, for example:
\begin{lstlisting}[language=coq]
(* a variable definition *)
Variables a b: nat.
(* a simple non-recursive function *)
Definition inc(a:nat) := a + 1.
(* axioms don't have to be proved *)
Axiom inc_ax: forall c:nat, inc(c) > c.
(* theorems rely on proving *)
Theorem inc_eq: forall c:nat, inc(c) = c + 1.
Proof.
(* interactive proving based on tactics *)
auto.
Qed.
\end{lstlisting}

As shown in this example, there are two rather different mode in Coq's interactive shell. When
we start Coq, we can write declarations and definitions in a functional-programming mode. Then, when
we start a \emph{Theorem}, or \emph{Lemma}, Coq jumps into the proving mode. We need to write
different \emph{tactics} to reduce the proving goal and finally finish the formal proof.

Furthermore, Coq is equipped with a set of well-written standard libraries. For example, as used in
this paper, \emph{Stream} provides a co-inductive
definition of infinite lists, and \emph{Reals} defines various operations and theorems on real
numbers. Usually, quite a few lemmas and theorems are pre-defined in such libraries, making it
substantially easier to prove our goals.

@@

Z3 \cite{MouraB08} is an efficient SMT (Satisfiability Modulo Theories) solver freely available from Microsoft. It can be used in various software verification and analysis applications. Acting as a SMT solver, Z3 expands to deciding the satisfiability (or dually the validity) of first order formulas with respect to combinations of theories such as: arithmetic, bit-vectors, arrays, and uninterpreted functions. Given the data and time constraints of connectors, it allows us to decide the satisfiability against properties or refinement relations. Z3 provides bindings for several programming languages. In this paper, we use \emph{Z3 python-bindings} to construct the models and carry out experiments. The following example provides an intuitive understanding of the Z3 solver.
\begin{lstlisting}
x, y = Int('x'), Int('y')

s = Solver()
s.add(x > 10, Or(x + y > 3, x - y < 2))
# Solving constraints in the solver s
print s.check()

# Create a new scope
s.push()
s.add(y < 11)
# Solving updated set of constraints
print s.check()

# Restoring state
s.pop()
# Solving restored set of constraints
print s.check()
\end{lstlisting}

This example contains two constraints at first. From these two constraints, we can see that Z3 supports Boolean operators, such as \emph{And}, \emph{Or}, \emph{Not} and \emph{Implies}, apart from the operators like $<$, $>$ for comparison.
The function \emph{Solver} creates a general purpose solver instance where constraints can be asserted through its \emph{add} method. The method \emph{check} solves the asserted constraints. Finally, the result is satisfiable (unsatisfiable) if a solution is (not) found, respectively. A solver may fail to solve a system of constraints and \emph{unknown} is returned. Each solver maintains a stack of assertions. The command \emph{push} creates a new scope by saving the current stack size. The command \emph{pop} removes all constraints that have been asserted after the last \emph{push} command.

\section{Formal Modeling in Coq}\label{sec:channelandoperator}
In this section, we show how primitive connectors, i.e., channels, and operators for connector composition are specified in Coq and used for modeling of complex connectors. Basic channels, which can be regarded as axioms of the whole framework, are specified as logical predicates illustrating the relation between the timed data streams of input and output. When we need to construct a more complex connector, appropriate composition operators are applied depending on the topological structure of the connector.

@@

%\subsection{Timed data streams}
The behavior of a connector can be formalized by means of data-flows at its sink and source nodes which are essentially infinite sequences. With the help of the stream library in Coq, such infinite data-flows can be defined as \emph{timed data streams}:
\begin{lstlisting}[language=coq]
Definition Time := R.
Definition Data := nat.
(*Inductive Data : Set :=
    |Natdata : nat-> Data
    |Empty : Data.*)
Definition TD := Time * Data.
Variable Input : Stream TD.
Variable Output : Stream TD.
\end{lstlisting}
%
In our framework, time is represented by real numbers. %Benefit from the completeness of real number system, we can express and carry out the effective operation of a quantity at any precision request.
The continuity of the set of real numbers is sufficiently enough for our modeling approach. Also the
continuous time model is more appropriate than the discrete model since it is very expressive and closer to the nature of time in the real world. Thus, the time sequence consists of increasing and diverging time moments. For simplicity, here we take the natural numbers as the definition of data, which can be easily expanded according to different application domains. The Cartesian product of time and data defines a TD object.
We use the stream module in Coq to produce streams of TD objects.

%\subsection{Auxiliary functions and predicates}
Some auxiliary functions and predicates are defined to facilitate the representation of axioms for basic channels in Reo. This part can be extended for further use in different problems.
The terms \emph{PrL} and \emph{PrR} take a pair of values (a, b) that has Cartesian product type A $\times$ B as the argument and return the first or second value of the pair, respectively.
%\begin{verbatim}
%    Definition PrL {A B: Type}(pair: A * B) := (...).
%    Definition PrR {A B: Type}(pair: A * B) := (...).
%\end{verbatim}
The following functions provide some judgment of time, which can make
the description of axioms and theorems for connectors more concise and
clear: \emph{Teq} means that time of two streams are equal and \emph{Tneq} has the
opposite meaning. \emph{Tle}  (\emph{Tgt}) represents that time of the first stream is
strictly less (greater) than the second stream. The judgement about
equality of data is analogous to the judgement of time. The complete
definition of these functions can be found at \cite{reo2coq2Z3}.

@@

We use a pair of predicates to describe the constraints on time and data, respectively, and their intersection to provide the complete specification of basic channels. This model offers convenience for the analysis and proof of connector properties. In the following, we present a few examples of the formal model of basic channels.

@@

The channel types FIFO and FIFO$n$ where $n$ is an integer greater than $0$ represent the
typical unbounded and bounded asynchronous FIFO channels. A write to a FIFO channel always succeeds, and a write to a FIFO$n$ channel succeeds only if the
number of data items in its buffer is less than its bounded capacity $n$. A read or take operation on a FIFO or FIFO$n$ channel suspends until the first data item
in the channel buffer can be obtained and then the operation succeeds.
%We use the $\overrightarrow{\emph{a}}$ to denote the tail of  a sequence \emph{a}.
For simplicity, we take the FIFO1 channel as an example. This channel type requires that the time when it consumes a data item through its source end is earlier than the time when the data item is delivered through its sink end. Besides, as the buffer has the capacity 1, time of the next data item that flows in should be later than the time when the data in the buffer is delivered. We use intersection of predicates in its definition as follows:
\begin{lstlisting}[language=coq]
Definition FIFO1(Input Output:Stream TD) : Prop :=
    Tle Input Output /\ Tle Output (tl Input) /\ Deq Input Output.
\end{lstlisting}

For a FIFO1 channel whose buffer already contains a data element \emph{e}, the communication can be initiated only if the data element \emph{e} can be taken via the sink end. In this case, the data stream that flows out of the channel should get an extra element \emph{e} settled at the beginning of the stream. And time of the stream that flows into the channel should be earlier than time of the tail of the stream that flows out. But as the buffer contains the data element \emph{e}, new data can be written into the channel only after the element \emph{e} has been taken. Therefore, time of the stream that flows out is earlier than time of the stream that flows in. The channel can be represented as:
\begin{lstlisting}[language=coq]
Definition FIFO1e(Input Output:Stream TD)(e:Data) : Prop :=
    Tgt Input Output /\ Tle Input (tl Output)
    /\ PrR (hd Output) = e  /\ Deq Input (tl Output).
\end{lstlisting}

In the following we choose Axiom to define LossySync
because it is easier to use the coinductive expression to specify
its behavior.

@@

Defining basic channels by intersection of predicates provides the following benefits:
\begin{itemize}
\item Firstly, this makes the model intuitive and concise as each predicate describes a simple order relation on time or data.
\item Secondly, we can easily split predicates for proofs of different properties which can make the proving process simpler.
\end{itemize}
%Each predicate can be viewed as an atom. The theorems and propositions which will be described later are also illustrated by atomic expressions.

>>>>

%\usetikzlibrary{external}
%\tikzexternalize[prefix=figures/]
%\tikzset{external/force remake}

%\newcommand{\redt}[1]{\textcolor{red}{#1}}
\newcommand{\liyi}[1]{\textcolor{blue}{#1}}
\newcommand{\xy}[1]{{#1}}

@@

%\maketitle
\begin{abstract}
Reo is a channel-based exogenous coordination language in which complex coordinators, called connectors, are compositionally built out of simpler ones. In this paper, we present an approach to model and reason about connectors using Coq and Z3.
Both models reflect the original structure of connectors \liyi{as closely as possible}. In our framework, both basic connectors (channels) and composition operations are \xy{modeled as axioms.} Furthermore, complex connectors are modeled as the combination of logical predicates which correspond to simpler connectors. With such definitions provided, connector properties, as well as equivalence and refinement relations between different connectors, can be formalized as \emph{goals} in Coq and proved using pre-defined \emph{tactics}, if satisfied by connectors. \xy{When failing to prove whether a property is satisfiable or not with Coq, we use Z3, an SMT solver, to search for possible bounded counter examples automatically.}
% Particularly, to fulfill the proof of refinement relation, Z3 is used as an assistant tool for solving logical formulas.
%\paragraph*{Keywords:}

@@

Modern software systems are typically distributed over large networks of \xy{computing devices.} \xy{Usually the components}
that comprise a system do not exactly fit together as pieces of a jigsaw puzzle, but leave significant interfacing gaps
that must somehow be filled with additional code. Compositional coordination models and languages provide a formalization
of the ``glue code" that interconnects the constituent \liyi{components, organizes} the mutual interactions \xy{between them} in a
distributed processing environment, and \xy{plays} a crucial role for the success of component-based systems in the past decades.

As an example,  \xy{Reo \cite{Arb04} is a channel-based exogenous coordination language, that offers
a powerful framework for implementation of coordinating component connectors.} Connectors provide the protocols that control and organize
the communication, synchronization and cooperation among the components that they interconnect. Primitive connectors, called {\em channels}
in Reo, can be composed to build complex connectors. Reo has been successfully applied in different application domains, such as
service-oriented computing and bioinformatics \cite{CCA04, SA07}. \xy{In recent years,
verifying the correctness of connectors is becoming a critical challenge, especially due to the advent of Cloud computing
technologies. The rapid \liyi{growth in size} and complexity of the computing infrastructures leads to more complex structure of connectors, makes it more difficult to model
and verify connector properties, and thus decreases the confidence on the correctness of connectors.}

Several works have been done for formal modeling and \xy{verification of connectors.} An operational semantics for Reo using Constraint Automata
(CA) was provided by Baier et al. \cite{BSAR06}, and later the symbolic model checker Vereofy \cite{BBK+10} was developed,
which can be used to check CTL-like properties.
% However, modeling unbounded primitives or even bounded primitives with unbounded
% data domains is impossible with finite constraint automata. Bounded large data domains cause an
% explosion in the constraint automata model which becomes problematic.
% A model for Reo connectors based on the idea of coloring a connector with possible data flows to
% resolve synchronization and exclusion constraints was presented by Clarke et al. \cite{CCA07}. Unlike the operational semantics,
% data sensitive behavior, which is supported by Reo, are not captured by the coloring approach.
Besides, one attractive approach is \xy{to translate Reo} to other formal models such as Alloy
\cite{KSA+08}, mCRL2 \cite{KKV12}, UTP \cite{AAA+09,SAA+12}, etc.,
which makes it possible to take advantage of existing verification
tools. A comparison of existing semantic models for Reo can be found
in \cite{JA12}.

In a previous paper \cite{ZHL+17} we showed how to formally model and reason about connectors' properties in Coq. The basic idea of our approach is to model the behavior of \xy{a connector as a logical predicate} which describes the relation among the timed data streams on the input and output nodes, and to reason about connectors' properties, as well as the equivalence and refinement relations between connectors, by using proof principles and tactics in Coq. \xy{Almost all existing approaches for modeling and verification of Reo connectors are state-based or automata-based, in which the automata for complex connectors are constructed through product operation on simpler ones. As a result, the number of states of connectors  may grow exponentially, which makes it inefficient and even impossible to verify connector properties in model checkers. Compared with existing approaches for verifying connectors' properties \cite{BBK+10,KB09,KKV12}, reasoning about connectors using theorem provers like Coq is especially helpful when we take infinite behavior into consideration. The coinductive proof principle makes it possible to prove connectors' properties easily while it is difficult (sometimes impossible) for other approaches (like model checking) because of the huge (or maybe infinite) number of states.}

The approach in this paper is not a brand new idea, as we have already provided a solution for modeling Reo in Coq in \cite{LS15}, where connectors are represented in a constructive way, and verification is essentially based on simulations. We do believe that the approach in this paper is reasonably different from its predecessor \cite{LS15} where Coq seldom shows its real power. To be more specific, this work has its certain
advantages comparing with \cite{LS15} in the following aspects:
\begin{itemize}
\item {\bf Modeling Method:} We use axioms to describe basic channels and their composition operations, which is more natural on a proof-assistant platform than the simulation-based approach in \cite{LS15}.
\item {\bf Expressive power:} Any valid Coq expression can be used to depict properties, which is obviously more powerful than just using LTL formulas in \cite{LS15}. \liyi{Then}, \xy{support for continuous time behavior is also possible in our approach in this paper as we make use of real numbers to represent time values.}
\end{itemize}

\xy{
  This paper is an extension of out earlier work \cite{ZHL+17} where equivalence and refinement relations can be proved among different connectors, while the approach in \cite{LS15} is not capable of either equivalence or refinement checking.
  In the current paper, we aim to provide a more complete approach to formally model and reason about Reo connectors compared with the approach in \cite{ZHL+17}. The main drawback in \cite{ZHL+17} is that \liyi{sometimes the user fails} to prove that the complementary refinement relation holds or not using Coq. This problem is partially made up by using Z3 \cite{MouraB08} in this paper. While Coq provides a platform to prove properties with a series of tactics and strategies to be considered, Z3 can be used as a component that \liyi{can build automatically counter-examples when the properties do not hold}. The logical formulas involved in the \liyi{Reo models} should \liyi{reflect, in the same way as the Coq model, the} constraints of connectors and be consistent with their behavior.
  % Adopting the Z3 SMT solver in this work makes it possible to automatically find bounded counter examples when the refinement relations do not hold.
}

The paper is organized as follows. After this general introduction, we briefly summarize Reo, Coq and Z3
in Section \ref{sec:pre}. Section \ref{sec:channelandoperator} introduces the
formal model of basic channels, operators and complex connectors in Coq. Section
\ref{sec:verification} shows how to reason about connector properties and equivalence (or
refinement) relations under our Coq formalization. Section \ref{sec:refinement} provides a Reo model in Z3 and shows the usage of Z3 in refinement checking.
In Section \ref{sec:conclusion}, we conclude with some further research directions. Full source codes \xy{are provided} at \cite{reo2coq2Z3} for further reference.

@@

In this section, we provide a brief introduction to the coordination language Reo, \liyi{the Coq proof assistant} and the Z3 SMT solver.

\subsection{The Coordination Language Reo}
Reo is a channel-based exogenous coordination language \xy{where complex coordinators, called connectors,}
are compositionally built out of simpler ones \cite{Arb04}.
The simplest connectors are channels with well-defined behavior such as synchronous channels, FIFO channels, etc.
% Each channel in Reo has exactly two directed ends, with their own identities.
There are two types of channel ends: \emph{source} ends and \emph{sink} ends. A source channel end accepts
data into the channel, and a sink channel end dispenses data out of the channel.
\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{basics.pdf}
  \caption{Four types of basic channels.}\label{fig:basicchannel}
\end{figure}

The graphical notations of some basic channels are presented in Figure~\ref{fig:basicchannel}, and their behavior can be interpreted as follows:
\begin{itemize}
\item{\textbf{Sync:} a synchronous channel with one source end and one sink end. The pair of I/O operations on its two ends \xy{can only succeed simultaneously.}}
\item{\textbf{SyncDrain:} a synchronous channel which has two source ends.
 The pair of input operations on its two ends \xy{can only succeed} simultaneously.
 All data items written to this channel are lost.}
\item{\textbf{FIFO$n$:} an asynchronous channel with one source end and one sink end, and a bounded buffer with capacity $n$.
It can \xy{accept $n$ data items from its source end before emitting data on its sink end.} The accepted data items are kept in the internal buffer, and dispensed to
the sink end in FIFO order.
Especially, the FIFO$1$ channel is an instace of FIFO$n$ where the buffer capacity is 1.}
%\item{\textbf{AsyncDrain:}  an asynchronous channel} which has two source ends. The channel guarantees that the operations on its two ends never succeed simultaneously. All data items written to this channel are lost.
\item{\textbf{LossySync:} a synchronous channel with one source end
    and one sink end. The source end always accepts all data items. If
    there is no matching output operation on the sink end of the
    channel at the time that a data item is accepted, then the data
    item is lost; otherwise, the channel transfers the data item
    as a Sync channel, and the output operation at the sink end succeeds.}
\end{itemize}

@@

Complex connectors are constructed by composing simpler ones \xy{via the \emph{join} and \emph{hiding} operations.} Channels are joined together in nodes. The set of channel ends coincident on a node is disjointly partitioned into the sets of source and sink channel ends that coincide on the node, respectively. Nodes are categorized into source, sink and mixed nodes, depending on whether all channel ends that coincide on a node are source ends, sink ends or a combination of the two. The hiding operation is used to hide the internal topology of a connector. The hidden nodes can no longer be accessed or observed from outside. There are three types of operations for channel composition: \emph{flow-through}, \emph{merge} and \emph{replicate}. Figure~\ref{fig:channelcomposition} provides the graphical representation of these operations. \xy{Further details about Reo and its semantics can be found in \cite{Arb04, AR03, BSAR06}.}

@@

Coq \cite{huet1997coq} is a widely-used \xy{proof assistant, where denotational formalizations (e.g.
theorem and hypothesis) and operational ones} (e.g. functions and algorithms) are naturally
integrated. Moreover, it allows the interactive construction of formal proofs.
The formal language used in Coq, \emph{Gallina}, provides a convenient way to define
both programming statements and mathematical propositions, for example:
\begin{lstlisting}[language=coq]
(* a variable definition *)
Variables a b: nat.
(* a simple non-recursive function *)
Definition inc(a:nat) := a + 1.
(* axioms don't have to be proved *)
Axiom inc_ax: forall c:nat, inc(c) > c.
(* theorems rely on proving *)
Theorem inc_eq: forall c:nat, inc(c) = c + 1.
Proof.
(* interactive proving based on tactics *)
auto.
Qed.
\end{lstlisting}
As shown in this example, \xy{there are two different modes} in Coq's interactive shell. When
we start Coq, we can write declarations and definitions in a functional-programming mode. Then, when
we start a \emph{Theorem}, or \emph{Lemma}, Coq jumps into the proving mode. We need to \xy{write
\emph{tactics} }to reduce the proving goal and finally finish the formal proof.

\xy{Coq is equipped with a set of well-written standard libraries. }For example, as used in
this paper, \emph{Stream} provides a co-inductive
definition of infinite lists, and \emph{Reals} defines various operations and theorems on real
numbers. Usually, quite a few lemmas and theorems are pre-defined in such libraries, making it
substantially easier to prove our goals.

@@

Z3 \cite{MouraB08} is an efficient SMT (Satisfiability Modulo Theories) solver freely available from Microsoft. It \liyi{has been} used in various software verification and analysis applications. \xy{Z3 expands to deciding the satisfiability }(or dually the validity) of first order formulas with respect to combinations of theories such as: arithmetic, bit-vectors, arrays, and uninterpreted functions. Given the data and time constraints of connectors, \xy{it allows us to verify the satisfiability \liyi{of} properties or refinement relations.} Z3 provides bindings for several programming languages. In this paper, we use \emph{Z3 python-bindings} to construct the models and carry out experiments. The following example provides an intuitive understanding of the Z3 solver.
\begin{lstlisting}
x, y = Int('x'), Int('y')

s = Solver()
s.add(x > 10, Or(x + y > 3, x - y < 2))
# Checking satisfiability of constraints in the solver s
print s.check()

# Create a new scope
s.push()
s.add(y < 11)
# Checking satisfiability of updated set of constraints
print s.check()

# Restoring state
s.pop()
# Checking satisfiability of restored set of constraints
print s.check()
\end{lstlisting}

This example contains two constraints at first. From these two constraints, \xy{we can see that Z3 supports operators like $<$, $>$ for comparison, apart from boolean operators, such as \emph{And}, \emph{Or}, \emph{Not} and \emph{Implies}.}
The function \emph{Solver} creates a general purpose solver instance where constraints can be asserted through its \emph{add} method. The method \emph{check} \liyi{assesses the satisfiability of the} asserted constraints. \xy{Finally, the result is satisfiable (unsatisfiable) if the solver returns \emph{sat} (\emph{unsat}) respectively.} A solver may fail to \liyi{check the satisfiability of} a system of constraints and \emph{unknown} is returned. Each solver maintains a stack of assertions. The command \emph{push} creates a new scope by saving the current stack size. The command \emph{pop} removes all constraints that have been asserted after the last \emph{push} command.

\section{Formal Modeling in Coq}\label{sec:channelandoperator}
In this section, we show how primitive connectors, i.e., channels, and operators for connector composition are specified in Coq and used for \xy{modeling complex connectors. }Basic channels, which can be regarded as axioms of the whole framework, are specified as logical predicates illustrating the relation between the timed data streams of input and output. When we need to construct a more complex connector, appropriate composition operators are applied depending on the topological structure of the connector.

@@

%\subsection{Timed data streams}
The behavior of a connector can be formalized by means of data-flows at its sink and source nodes which are essentially infinite sequences. With the help of the stream library in Coq, such infinite data-flows can be defined as \emph{timed data streams}:
\begin{lstlisting}[language=coq]
Definition Time := R.
Definition Data := nat.
Definition TD := Time * Data.
Variable Input : Stream TD.
Variable Output : Stream TD.
\end{lstlisting}
%
In our framework, time is represented by real numbers. %Benefit from the completeness of real number system, we can express and carry out the effective operation of a quantity at any precision request.
The continuity of the set of real numbers is sufficiently enough for our modeling approach. Also the
continuous time model is more appropriate than the discrete model since it is very expressive and closer to the nature of time in the real world. \xy{Thus, the time sequence consists of increasing and diverging time moments.} For simplicity, here we take the natural numbers as the definition of data, which can be easily expanded according to different application domains. The Cartesian product of time and data defines a TD object.
We use the stream module in Coq to produce streams of TD objects.

%\subsection{Auxiliary functions and predicates}
Some auxiliary functions and predicates are defined to \liyi{ease} the representation of axioms for basic channels in Reo. This part can be extended for further use in different problems.
The terms \emph{PrL} and \emph{PrR} take a pair of values (a, b) that has Cartesian product type A $\times$ B as the argument and return the first or second value of the pair, respectively.
%\begin{verbatim}
%    Definition PrL {A B: Type}(pair: A * B) := (...).
%    Definition PrR {A B: Type}(pair: A * B) := (...).
%\end{verbatim}
The following functions provide some judgment of time, which can make
the description of axioms and theorems for connectors more concise and
clear: \emph{Teq} \xy{means that values of time in \liyi{the two streams taken as parameters} are equal} and \emph{Tneq} has the
opposite meaning. \emph{Tle}  (\liyi{resp.} \emph{Tgt}) represents that time of the first stream is
strictly \liyi{lesser} (greater) than the second stream. The judgement about
equality of data is analogous to the judgement of time. \xy{\emph{Deq} represents that the data of two streams are equal.}The complete
definition of these functions can be found at \cite{reo2coq2Z3}.

@@

\xy{We use the previous predicates} to describe the constraints on time and data, respectively, \xy{and their conjunction} to provide the complete specification of basic channels. \xy{This approach provides a relational semantics for Reo using predicates that relates the inputs and outputs of connectors.} This model offers convenience for the analysis and proof of connector properties. In the following, we present a few examples of the formal model of basic channels.

@@

The channel types FIFO and FIFO$n$ where $n$ is an integer \liyi{strictly greater} than $0$ represent the
typical unbounded and bounded asynchronous FIFO channels. A write to a FIFO channel always succeeds, and a write to a FIFO$n$ channel succeeds only if the
number of data items in its buffer is less than its bounded capacity $n$. A read or take operation on a FIFO or FIFO$n$ channel suspends until the first data item
in the channel buffer can be obtained and then the operation succeeds.
%We use the $\overrightarrow{\emph{a}}$ to denote the tail of  a sequence \emph{a}.
For simplicity, we take the FIFO1 channel as an example. This channel type requires that the time when it consumes a data item through its source end is \liyi{strictly earlier} than the time when the data item is delivered through its sink end. Besides, as the buffer has the capacity 1, time of the next data item that flows in should be \liyi{strictly later} than the time when the data in the buffer is delivered. We use \xy{the conjunction of predicates} in its definition as follows:
\begin{lstlisting}[language=coq]
Definition FIFO1(Input Output:Stream TD) : Prop :=
    Tle Input Output /\ Tle Output (tl Input) /\ Deq Input Output.
\end{lstlisting}

For a FIFO1 channel whose buffer already contains a data element \emph{e}, the communication can be initiated only if the data element \emph{e} can be taken via the sink end. In this case, the data stream that flows out of the channel should get an extra element \emph{e} settled at the beginning of the stream. And time of the stream that flows into the channel should be \liyi{strictly earlier} than time of the tail of the stream that flows out. But as the buffer contains the data element \emph{e}, new data can be written into the channel only after the element \emph{e} has been taken. Therefore, time of the stream that flows out is \liyi{strictly earlier} than time of the stream that flows in. The channel can be represented as:
\begin{lstlisting}[language=coq]
Definition FIFO1e(Input Output:Stream TD)(e:Data) : Prop :=
    Tgt Input Output /\ Tle Input (tl Output)
    /\ PrR (hd Output) = e  /\ Deq Input (tl Output).
\end{lstlisting}

In the following we \liyi{define LossySync as an Axiom}
because it is easier to use than the coinductive \liyi{expression that specifies its behavior}.

@@

\xy{Defining basic channels by conjunction and disjunction of predicates provides the following benefits:}
\begin{itemize}
\item Firstly, \liyi{this makes the structure of the definition in Coq identical to the usual implementation}.
\item Secondly, we can easily split predicates for proofs of different properties which can make the proving process simpler.
\end{itemize}
%Each predicate can be viewed as an atom. The theorems and propositions which will be described later are also illustrated by atomic expressions.
